A survey of generic IND-CCA2 transformations

### 2024-07-10 faster-fo-transform/cca-transform-survey/main.tex
Described kyber and ML-KEM's KEM construction

\begin{figure}[H]
    \center
    \begin{minipage}{0.33\textwidth}
        \begin{algorithm}[H]
            \caption{$\keygen^{\not\bot}$}\label{alg:u-notbot-keygen}
            \begin{algorithmic}[1]
                \State $(\pk, \sk^\prime) \leftsample \keygen(1^\lambda)$
                \State $s \leftsample \mathcal{M}_\pke$
                \State $\sk = (\sk^\prime, s)$
                \State \Return $(\pk, \sk)$
            \end{algorithmic}
        \end{algorithm}
    \end{minipage}
    \hfill
    \begin{minipage}{0.33\textwidth}
        \begin{algorithm}[H]
            \caption{$\encap^{\not\bot}(\pk)$}\label{alg:u-notbot-encap}
            \begin{algorithmic}[1]
                \State $m \leftsample \mathcal{M}_\pke$
                \State $c \leftsample \encrypt(\pk, m)$
                \State $K \leftarrow H(m, c)$
                \State \Return $(c, K)$
            \end{algorithmic}
        \end{algorithm}
    \end{minipage}
    \hfill
    \begin{minipage}{0.32\textwidth}
        \begin{algorithm}[H]
            \caption{$\decap^{\not\bot}(\sk, c)$}\label{alg:u-notbot-decap}
            \begin{algorithmic}[1]
                \State $\hat{m} \leftarrow \decrypt(\sk, c)$
                \If{$\hat{m} \in \mathcal{M}_\pke$}
                    \State $K \leftarrow H(\hat{m}, c)$
                    \State \Return $K$
                \EndIf
                \State \Return $H(s, c)$
            \end{algorithmic}
        \end{algorithm}
    \end{minipage}
    \caption{$U^{\not\bot}$ routines}
\end{figure}

@@

\subsubsection{KEM with implicit rejection from rigid PKE}
% TODO: write out this section

@@

>>>>

\begin{figure}[H]
    \center
    \begin{minipage}{0.33\textwidth}
        \begin{algorithm}[H]
            \caption{$\keygen^{\not\bot}$}\label{alg:u-notbot-keygen}
            \begin{algorithmic}[1]
                \State $(\pk, \sk^\prime) \leftsample \keygen(1^\lambda)$
                \State $s \leftsample \mathcal{M}_\pke$
                \State $\sk = (\sk^\prime, s)$
                \State \Return $(\pk, \sk)$
            \end{algorithmic}
        \end{algorithm}
    \end{minipage}
    \hfill
    \begin{minipage}{0.33\textwidth}
        \begin{algorithm}[H]
            \caption{$\encap^{\not\bot}(\pk)$}\label{alg:u-notbot-encap}
            \begin{algorithmic}[1]
                \State $m \leftsample \mathcal{M}_\pke$
                \State $c \leftsample \encrypt(\pk, m)$
                \State $K \leftarrow H(m, c)$
                \State \Return $(c, K)$
            \end{algorithmic}
        \end{algorithm}
    \end{minipage}
    \hfill
    \begin{minipage}{0.32\textwidth}
        \begin{algorithm}[H]
            \caption{$\decap^{\not\bot}(\sk, c)$}\label{alg:u-notbot-decap}
            \begin{algorithmic}[1]
                \State $\hat{m} \leftarrow \decrypt(\sk, c)$
                \If{$\hat{m} \in \mathcal{M}_\pke$}
                    \State $K \leftarrow H(\hat{m}, c)$
                    \State \Return $K$
                \EndIf
                \State \Return $H(s, c)$
            \end{algorithmic}
        \end{algorithm}
    \end{minipage}
    \caption{$U^{\not\bot}$ routines}\label{fig:u-notbot-routines}
\end{figure}

@@

\subsubsection{KEM with implicit rejection from rigid PKE}
\begin{figure}[H]
    \caption{$U^{\not\bot}_m$ routines}\label{fig:u-notbot-m-routines}
\end{figure}
% TODO: write out this section

@@

For constructing $\kem$ from $\pke$, Kyber uses the $U^{\not\bot}$ transformation while ML-KEM uses the $U^{\not\bot}_m$ transformation. In implementation, however, there are two deviations from algorithm listed in figure \ref{fig:u-notbot-routines} and \ref{fig:u-notbot-m-routines}.

The first deviation is the inclusion of the public key $\pk$ when deriving the coin. Instead of deriving coin from the plaintext only $r \leftarrow G(m)$, Kyber and ML-KEM derive the coin from both the plaintext and (a hash of) the public key $(K, r) \leftarrow G(m, H(\pk))$. The main rationale \cite{bos2018crystals} behind this design choice is to prevent adversaries from precomputing $m$ values that have a high likelihood of triggering decryption failure, which can reveal information about the secret key.

Recall from the decryption routine $\decrypt_\pke$:

\begin{equation*}
    \begin{aligned}
        \hat{m} &= \mathbf{s}^\intercal \cdot \mathbf{c}_1 - c_2 \\
        &= \mathbf{s}^\intercal \cdot (
            A^\intercal \cdot \mathbf{r}_1 + \mathbf{e}_1
        ) - (
            \mathbf{t}^\intercal \cdot \mathbf{r}_1 + e_2 
            + m \cdot \round{\frac{q}{2}}
        ) \\
        &= \mathbf{s}^\intercal \cdot (
            A^\intercal \cdot \mathbf{r}_1 + \mathbf{e}_1
        ) - (
            (
                A\cdot\mathbf{s} + \mathbf{e}
            )^\intercal \cdot \mathbf{r}_1 + e_2 
            + m \cdot \round{\frac{q}{2}}
        ) \\
        &= \mathbf{s}^\intercal \cdot \mathbf{e}_1 
            - \mathbf{e}^\intercal \cdot \mathbf{r}_1 
            + e_2 + m \round{\frac{q}{2}}
    \end{aligned}
\end{equation*}

A decryption failure occurs if the noise term $\mathbf{s}^\intercal\cdot\mathbf{e}_1 - \mathbf{e}^\intercal\cdot\mathbf{r}_1 + e_2$ has large coefficient, which roughly translates to $\mathbf{s}^\intercal\cdot\mathbf{e}_1$ having large coefficients. If the coin $r \leftarrow G(m)$ is entirely derived from the adversary, then the adversary can compute a large number of $\mathbf{r}_1, \mathbf{e}_1, e_2$ offline, then observe which set of $\mathbf{r}_1, \mathbf{e}_1, e_2$ trigger decryption failure, which then can be used to infer information about $\mathbf{s}$ and $\mathbf{e}$. On the other hand, if the coin is derived from both the message and the public key, then the adversary needs to compute a large set of value per keypair, which pushes this attack into infeasibility even for adversaries with large quantum computers.

A second deviation is that instead of deriving the shared secret from the plaintext and the ciphertext $K \leftarrow H(m, c)$, Kyber derives the shared secret from the plaintext and \emph{a hash of the ciphertext}. Since ML-KEM uses the $U^{\not\bot}_m$ transformation, this deviation does not apply. \cite{bos2018crystals} cites benchmarking for non-incremental hash APIs to be the reason for this deviation, which I do not fully understand, but has likely less to do with the security of the scheme and more with verification and compliance.

The details of Kyber and ML-KEM's KEM construction are described in figures \ref{fig:kyber-kem-routines} and \ref{fig:mlkem-kem-routines} respectively.

\begin{figure}[H]
    \begin{minipage}{0.3\textwidth}
        \begin{algorithm}[H]
            \caption{$\keygen_\monospace{KyberKEM}$}\label{alg:kyber-kem-keygen}
            \begin{algorithmic}[1]
                \State $(\pk_\pke, \sk_\pke) \leftsample \keygen_\pke()$
                \State $z \leftsample R_{0,1}$
                \State $h \leftarrow H(\pk_\pke)$
                \State $\sk_\kem \leftarrow (\sk_\pke, z, h, \pk_\pke)$
                \State $\pk_\kem \leftarrow \pk_\pke$
                \State \Return $(\pk_\kem, \sk_\kem)$
            \end{algorithmic}
        \end{algorithm}
    \end{minipage}
    \hfill
    \begin{minipage}{0.33\textwidth}
        \begin{algorithm}[H]
            \caption{$\encap_\monospace{KyberKEM}(\pk)$}\label{alg:kyber-kem-encap}
            \begin{algorithmic}[1]
                \State $m \leftsample R_2$
                \State $(K, r) \leftarrow G(H(\pk), m)$
                \State $c \leftarrow \encrypt_\pke(\pk, m; r)$
                \State $K \leftarrow \monospace{KDF}(K, c)$
                \State \Return $(c, K)$
            \end{algorithmic}
        \end{algorithm}
    \end{minipage}
    \hfill
    \begin{minipage}{0.33\textwidth}
        \begin{algorithm}[H]
            \caption{$\decap_\monospace{KyberKEM}(\sk, c)$}\label{alg:kyber-kem-decap}
            \begin{algorithmic}[1]
                \State $(\sk_\pke, z, h, \pk_\pke) \leftarrow \sk$
                \State $\hat{m} \leftarrow \decrypt_\pke(\sk_\pke, c)$
                \State $(\hat{K}, \hat{r}) \leftarrow G(h, \hat{m})$
                \State $\hat{c} \leftarrow \encrypt_\pke(\pk, \hat{m}; \hat{r})$
                \If{$\hat{c} = c$}
                    \State \Return $\monospace{KDF}(\hat{K}, H(\hat{c}))$
                \Else
                    \State \Return $\monospace{KDF}(z, H(\hat{c}))$
                \EndIf
            \end{algorithmic}
        \end{algorithm}
    \end{minipage}
    \caption{Kyber's KEM routines}\label{fig:kyber-kem-routines}
\end{figure}

\begin{figure}[H]
    \begin{minipage}{0.3\textwidth}
        \begin{algorithm}[H]
            \caption{$\keygen_\monospace{ML-KEM}$}\label{alg:mlkem-keygen}
            \begin{algorithmic}[1]
                \State $(\pk_\pke, \sk_\pke) \leftsample \keygen_\pke()$
                \State $z \leftsample R_2$
                \State $h \leftarrow H(\pke_\pke)$
                \State $\sk_\kem \leftarrow (\sk_\pke, z, h, \pk_\pke)$
                \State $\pk_\kem \leftarrow \pke_\pke$
                \State \Return $(\pk_\kem, \sk_\kem)$
            \end{algorithmic}
        \end{algorithm}
    \end{minipage}
    \hfill
    \begin{minipage}{0.33\textwidth}
        \begin{algorithm}[H]
            \caption{$\encap_\monospace{ML-KEM}(\pk_\kem)$}\label{alg:mlkem-encap}
            \begin{algorithmic}[1]
                \State $m \leftsample R_2$
                \State $(K, r) \leftarrow G(m, H(\pk))$
                \State $c \leftarrow \encrypt_\pke(\pk, m; r)$
                \State \Return $(c, K)$
            \end{algorithmic}
        \end{algorithm}
    \end{minipage}
    \hfill
    \begin{minipage}{0.33\textwidth}
        \begin{algorithm}[H]
            \caption{$\decap_\monospace{ML-KEM}(\sk, c)$}\label{alg:mlkem-decap}
            \begin{algorithmic}[1]
                \State $(\sk_\pke, z, h, \pk_\pke) \leftarrow \sk$
                \State $\hat{m} \leftarrow \decrypt_\pke(\sk_\pke, c)$
                \State $(\hat{K}, \hat{r}) \leftarrow G(\hat{m}, h)$
                \State $K^\prime \leftarrow \monospace{KDF}(z, c)$
                \State $\hat{c} \leftarrow \encrypt_\pke(\pk_\pke, \hat{m}; \hat{r})$
                \If{$\hat{c} = c$}
                    \State \Return $\hat{K}$
                \Else
                    \State \Return $K^\prime$
                \EndIf
            \end{algorithmic}
        \end{algorithm}
    \end{minipage}
    \caption{ML-KEM's KEM routines}\label{fig:mlkem-kem-routines}
\end{figure}
